---
title: "MB5370_Module_04_JC"
author: "James Coon"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Module 4, Data Science in R

### Load Packages
```{r Load Packages}
install.packages("tidyverse", 'ggplot2')
library(tidyverse, ggplot2)
```

### Load Data
```{r mpg}
mpg
```

### Create first plot
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

### Graphing Template:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(mapping = aes(<MAPPINGS>))

```{r Graphing}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, colour = class)) #color by class!

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class)) #point size by class!

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class)) #alpha is transparency, thus transparency by class!

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class)) #point shape by class!

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") #can manually make all points whatever color you want!
```
What happens if aes(colour = displ < 5)?
It creates a legend which displays values over a 5 dspl in turquoise and under in red.
```{r Question}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = displ < 5))
```

### Facets and Panel Plots

Creates plots in a grid format, where it's easier to compare them.
```{r Facets and Panel Plots}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2) #Creates 2 rows, regardless of amount of variables/plots

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(. ~ cyl) #If showing more than 1 variable as a row, use facet_grid instead of facet_wrap.

```
Exercise-

What does ncol and nrow do?
They choose the number of columns and rows respectively.

What other options control the layout of the individual panels? 
Reorder and as.table

## Fitting Simple Lines

```{r Fitting Simple Lines}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) #Plot as points

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy)) #Plot as smooth line! geom_smooth

ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv)) #Can change plot line by class!

ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv, color=drv, linetype = drv)) #Change some aesthetic based on drv

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy)) #Plots both line and points. Must just be used additively

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth() #Same as above, but more efficient

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth() #Style each point individually based upon many variables
```

#### Exercise

1. What geom would you use to draw a line chart? A boxplot? A histogram? An area chart?
line cart- geom_smooth
boxplot- geom_boxplot
histogram- geom_histogram or geom_bar

2. Run this code in your head and predict what the output will look like. Then, run the code in R and check your predictions.
```{r Exercise}
#I predict both graphs will be the same.

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()

ggplot() + 
  geom_point(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_smooth(data = mpg, mapping = aes(x = displ, y = hwy))
```

3. Will these two graphs look different? Why/why not?
No, because the specifications made in the second plot are listed in the first plot.  The first plot is simply more efficient because the specifications only have to be typed once.

## Transformations and Stats

```{r}
ggplot(data = diamonds) + #data found in ggplot2
  geom_bar(mapping = aes(x = cut))

ggplot(data = diamonds) + 
  stat_count(mapping = aes(x = cut)) #can often use stat and geom interchangeably
```
#### Overriding Defaults
```{r Overriding defaults}
demo <- tribble( #tribble is basically a dataframe
  ~cut,         ~freq,
  "Fair",       1610,
  "Good",       4906,
  "Very Good",  12082,
  "Premium",    13791,
  "Ideal",      21551
)
demo

ggplot(data = demo) +
  geom_bar(mapping = aes(x = cut, y = freq), stat = 'identity') #default stat is count

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, y = stat(prop), group = 1)) # y is a proportion rather than raw count
  
```

#### Plotting Statistical Details

Showing statistical summaries is a good ideal for transparency of your work
```{r}

ggplot(data = diamonds) + 
  stat_summary(  #Shows error bars when specified
    mapping = aes(x = cut, y = depth),
    fun.min = min,
    fun.max = max,
    fun = median)
```

## Aesthetic Adjustments

You can use commands such as color or fill to customize bars.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut)) #only outline is colored

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut)) #Fills bars

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity)) #Stacks clarity coloration into existing bars!

ggplot(data = diamonds) +
(mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity") #Alpha alters transparency

ggplot(data = diamonds) + 
  (mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity") #Transparent bars and colored outline.

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill") #Similar to stacking, but makes all the bars the same height.

ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge") #Dodge puts overlapping objects next to eachother instead of overlapping.

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), position = "jitter") #Jitter makes random noise in the plot to slightly and randomly space data out and make it mroe visible.
```

##The layered grammar of graphics

An updated plot template:

ggplot(data = <DATA>) + 
  <GEOM_FUNCTION>(
     mapping = aes(<MAPPINGS>),
     stat = <STAT>, 
     position = <POSITION>
  ) +
  <FACET_FUNCTION>

## Workshop 1 Assignment

IMPORT DATASET
```{r}
library(readxl)
API_ER_FSH_AQUA_MT_DS2_en_excel_v2_967 <- read_excel("C:/Users/jacoo/Downloads/API_ER.FSH.AQUA.MT_DS2_en_excel_v2_967.xls", 
    sheet = "Sheet2")
View(API_ER_FSH_AQUA_MT_DS2_en_excel_v2_967)
JA_Fisheries <- API_ER_FSH_AQUA_MT_DS2_en_excel_v2_967
```

Alter the plot!
```{r Assignment exercise}
JA_Fisheries$Year<- as.numeric(JA_Fisheries$Year) #This held us up for a half hour lol.  Need to assign to the data sheet, not just as.numeric(data)
JA_Fisheries <- na.omit(JA_Fisheries)
str(JA_Fisheries$Year)
ggplot(data= JA_Fisheries) + 
 geom_point(mapping = aes(x= Year, y= Japan), color="black") +
 geom_point(mapping = aes(x= Year, y= Australia), color="grey") +
 geom_smooth(aes(x= Year, y= Japan), color="black", method='loess') +
  geom_smooth(aes(x= Year, y= Australia), color="grey", method='loess') +
labs(y = "Aquaculture Production (Metric Tons)", x = "Year", color = 'Country') #This coloration is better, but not for color blindness.
```

```{r}
ggplot(data = JA_Fisheries) + 
  geom_point(aes(x = Year, y = Japan, color = "Japan", shape = "Japan")) +
  geom_point(aes(x = Year, y = Australia, color = "Australia", shape = "Australia")) +
  geom_smooth(aes(x = Year, y = Japan, color = "Japan"), method = "loess") +
  geom_smooth(aes(x = Year, y = Australia, color = "Australia"), method = "loess") +
  labs(y = "Aquaculture Production (tons)", x = "Year", color = "Country", shape= "Country")  # Legend title, and symbology is much better for the color blind, though color may not be ideal.
```
Caption- The aquaculture production in Australia is much lower than in Japan.  However, there is a noticeable increase in Australia's production recently (the 1990s) while Japan has been declining consistently since about 1995.  Data provided by World Data Bank.

#Workshop 2- GGplot2 for Communication

## Labels

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
    title = "Fuel efficiency generally decreases with engine size", #Main title label
    subtitle = "Two seaters (sports cars) are an exception because of their light weight", #smaller title label
    caption = "Data from fueleconomy.gov" #by default appears on the bottom of page
  )


ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_smooth(se = FALSE) +
  labs(
    x = "Engine displacement (L)", #axis titles
    y = "Highway fuel economy (mpg)",
    colour = "Car type" #legend title
  )
```

## Annotations

```{r}
best_in_class <- mpg %>%
  group_by(class) %>%
  filter(row_number(desc(hwy)) == 1) #This bit of code filters out the best mpg car for each car class!

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  geom_text(aes(label = model), data = best_in_class) #annotates each best car model for the mpg we selected earlier!  Annotation overlapping can be resolved with nudge()
```

## Scales

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) #Automatically selects scales for you, but you can tweak them if you need


ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class)) +
  scale_x_continuous() +
  scale_y_continuous() +
  scale_colour_discrete() #Can also provide ylims and xlims!
```

## Axis Ticks


```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_y_continuous(breaks = seq(15, 40, by = 5)) #Seq sets limits and selects ticks by the interval selected within the limits.

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  scale_x_continuous(labels = NULL) +
  scale_y_continuous(labels = NULL) #Set to null to rid of labels all together!
```

## Legends and Color Schemes

```{r}
base <- ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(colour = class))

base + theme(legend.position = "left")
base + theme(legend.position = "top")
base + theme(legend.position = "bottom")
base + theme(legend.position = "right") #Simply changes the position of the legend in relation to the plot
```

## Replacing a Scale

```{r}
ggplot(diamonds, aes(carat, price)) +
  geom_bin2d() + 
  scale_x_log10() + 
  scale_y_log10()  #Scale transformations can show important information!  Remember to say they were transformed for transparency.
```

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) #default color scale

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv)) +
  scale_colour_brewer(palette = "Set1") #Creates own color palette!

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = drv, shape = drv)) + #adding shape makes graph interpretable when black and white as well! good for color blindness and cheap printing.
  scale_colour_brewer(palette = "Set1")
```
```{r}
presidential %>%
  mutate(id = 33 + row_number()) %>%
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) #manually selects colors if needed to be pre defined
```
```{r}
#install.packages('viridis') #Sweet packages
#install.packages('hexbin') #Another one
library(viridis)
library(hexbin)

df <- tibble( # note we're just making a fake dataset so we can plot it
  x = rnorm(10000),
  y = rnorm(10000)
)
ggplot(df, aes(x, y)) +
  geom_hex() + # a new geom!
  coord_fixed()

ggplot(df, aes(x, y)) +
  geom_hex() +
  viridis::scale_fill_viridis() +
  coord_fixed()
```

## Themes

```{r}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_bw() #black and white theme

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_light() #lighter graph borders

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_classic() #no grid lines, only axis borders

ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  theme_dark() #Darkened graph background, difficult to read nless lighter data points

theme (panel.border = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.minor.y = element_blank(),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=8),
        panel.grid.major = element_blank(),
        legend.key = element_blank(),
        legend.background = element_blank(),
        axis.text.y=element_text(colour="black"),
        axis.text.x=element_text(colour="black"),
        text=element_text(family="Arial"))  #Can develop yuor own themes by setting every variable yourself
```

## Save and Export Your Figures

```{r}
ggplot(mpg, aes(displ, hwy)) + geom_point()

ggsave("my-plot.pdf")
#> Saving 7 x 4.32 in image to your working directory
```

##QFish Assignment

```{r Import Otter Trawl Data}
library(readxl)
export <- read_excel("C:/Users/jacoo/Downloads/export.xlsx")
View(export) #Import Dataset
QFish <- export
```

```{r Cleaning Data}
#Data is fairly clean, need to remove last year (2024) because it is incomplete.
#QFish <- QFish %>%
  #slice(-n())
#view(QFish)
#commented out as to not over-use the command and lose data
```

```{r Plotting}
QFish$Tonnes <- as.numeric(QFish$Tonnes)
QFish$Days <- as.numeric(QFish$Days)
QFish$CalendarYear <- as.numeric(QFish$CalendarYear)
ggplot(data= QFish) + 
  geom_point(aes(x = CalendarYear, y = Days, color = "Days"), alpha = 0.7) +
  geom_point(aes(x = CalendarYear, y = Tonnes, color = "Tonnes"), alpha = 0.7) +
  geom_smooth(aes(x = CalendarYear, y = Days, color = "Days"), method = 'loess', se= FALSE) +
  geom_smooth(aes(x = CalendarYear, y = Tonnes, color = "Tonnes"), method = 'loess', se= FALSE) +
  scale_color_manual(values = c("black", "red")) +
  labs(y = "Fishing Effort (Days) or Yield (Tonnes)", x = "Year", color = "Measure") +
  theme_bw() +
  theme(panel.grid = element_blank()) #Use minimal theme for clarity
```
Caption- Otter trawl efforts have decreased quite quickly since the 1990s while yields have been decreasing very slightly ever since. This could be due to an increase in technology efficiency and overfishing depleted populations. (Data from QFish.fisheries.qld.gov)






